Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-06-19 paper/main.tex
generalize spectral norm moment bound

\begin{lemma}{Moment of spectral norm of non-standard Gaussian random matrix}{ose}
    The matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m) \in \mathbb{R}^{m \times m}$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
        \leq  \sqrt{k + p} \left( 3 \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
    \end{equation}
    %\begin{equation}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
    %\end{equation}
    %If we choose 
\end{lemma}

\begin{proof}
    First, consider the above moment of a vector $\vct{\omega} \in \mathbb{R}^{m}$. We can rewrite
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        = \mathbb{E}\left[  \left( \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \mathbb{E}\left[  \left( \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \sqrt{ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] }.
        \label{equ:matvec-spectral-norm-simplification}
    \end{equation}
    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
    \begin{equation}
        \underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}, \underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}, \dots, \underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}.
    \end{equation}
    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
    \begin{equation}
        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
        \quad \text{and} \quad
        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell,
    \end{equation}
    and therefore
    \begin{equation}
        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
        \label{equ:singular-value-group-bound}
    \end{equation}
    This allows us to bound
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
        &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] && \text{(separate sum into groups)} \notag \\
        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$)} \notag \\
        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} && \text{(\refequ{equ:singular-value-group-bound} and $\chi_k^2$ moments \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + p). && \text{(\reflem{lem:gamma})}
    \end{align}
    Thus, together with \refequ{equ:matvec-spectral-norm-simplification} and using $\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$ for $a, b \geq 0$, we obtain
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        \leq \sqrt{k + p} \left( \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
        \label{equ:matvec-spectral-norm-moment-bound}
    \end{equation}

@@

    We can apply \reflem{lem:ose} with $\mtx{\Sigma} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with

>>>>

\begin{lemma}{Moment of spectral norm of non-standard Gaussian random matrix}{ose}
    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right]
        \leq  \sqrt{k + p} \left( 3 \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
    \end{equation}
    %\begin{equation}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
    %\end{equation}
    %If we choose 
\end{lemma}

\begin{proof}
    With the singular value decomposition $\mtx{A} = \mtx{U} \mtx{\Sigma} \mtx{V}^{\top}$ and the unitary invariance of the spectral norm, it can be seen that $\lVert \mtx{A} \mtx{\Omega} \rVert _2 = \lVert \mtx{\Sigma} \widetilde{\mtx{\Omega}} \rVert _2$ where $\widetilde{\mtx{\Omega}}$ is again a standard Gaussian random matrix. Hence, it is enough to bound $\mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]$ for a diagonal matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m)$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$.

    First, consider the above moment of a vector $\vct{\omega} \in \mathbb{R}^{m}$. We can rewrite
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        = \mathbb{E}\left[  \left( \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \mathbb{E}\left[  \left( \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \sqrt{ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] }.
        \label{equ:matvec-spectral-norm-simplification}
    \end{equation}
    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
    \begin{equation}
        \underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}, \underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}, \dots, \underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}.
    \end{equation}
    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
    \begin{equation}
        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
        \quad \text{and} \quad
        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell,
    \end{equation}
    and therefore
    \begin{equation}
        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
        \label{equ:singular-value-group-bound}
    \end{equation}
    This allows us to bound
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
        &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] && \text{(separate sum into groups)} \notag \\
        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$)} \notag \\
        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} && \text{(\refequ{equ:singular-value-group-bound} and $\chi_k^2$ moments \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + p). && \text{(\reflem{lem:gamma})}
    \end{align}
    Thus, together with \refequ{equ:matvec-spectral-norm-simplification} and using $\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$ for $a, b \geq 0$, we obtain
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        \leq \sqrt{k + p} \left( \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
        \label{equ:matvec-spectral-norm-moment-bound}
    \end{equation}

@@

    We can apply \reflem{lem:ose} with $\mtx{A} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with
